Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests

### 2013-11-06 paper.tex
ENH:  initial draft?

\title{Multivariate Neuroanalysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}

@@

ANTs (Advanced Normalization Tools) originated with the open source availing
of high-performance registration algorithms for neuroimage analysis
\citep{avants2008a} built upon the mature and well-vetted Insight Toolkit (ITK)
of the National Institutes of Health.  Since then, ANTs has grown to include 
several algorithmic solutions for robust medical image analysis including bias 
correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
\citep{avants2011}, template construction \citep{avants2010}, and cortical 
thickness estimation \citep{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').  
However, in the evolution of the toolkit, it became clear 
that robust statistical machinery was lacking for making inferences regarding
the data produced during the course of ANTs processing.  ANTsR was developed
specifically to provide an interface between ANTs, a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations, and the R project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate neuroimage analysis.

In addition to describing ANTsR basics and how it can be
used generally in multivariate neuroimaging studies, we showcase its use in
the particularly difficult application of supervised brain tumor
segmentation \cite{bauer2013} for which we use the widely popular 
random forests machine learning technique.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011}), 
we demonstrate that ANTsR significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework has proven extremely in 
actual practice, specifically
 in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

@@

Careful consideration of available options
led to the adoption of R to complement ANTs capabilities resulting in the
ANTsR package.
The R project for statistical computing,%
\footnote{
www.r-project.org
}
 or more compactly
`R', is an environment for statistical computation
and data visualization.  R's open source code base
and add-on packages coupled with its rapidly growing 
community of developers and users has caused wide-scale
adoption within both academia and industry.

@@

As mentioned earlier, we have made this entire framework
available as open source.  In addition to the ANTsR repository
already on github which houses both ANTs and ANTsR functionality, 
we created a special github repository specifically for this work
containing figures, references, and text.%
\footnote{
https://github.com/ntustison/ANTsAndArboles
}
Also, we posted all
scripts (R, shell, and perl) used to coordinate the ANTsR processing 
including (along with a brief description of each script):
\begin{itemize}
  \item {\tt applyModel.R}:  applies a random forest model to a new 
  feature data set from a testing subject resulting in a set of probability
  images (one for each label).
  \item {\tt applyTumorSegmentationModel.sh}:  generates the new feature image set 
  from the testing MRI (by calling {\tt createFeatureImages.sh}).
  organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
  and applies the random forest model using {\tt applyModel.R}. 
  \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
  segmentation on the computational cluster for a given cohort.
  \item {\tt createCSVFileFromModel.R}:  organizes the set of feature image
  file names in a csv file for input into {\tt applyModel.R}.
  \item {\tt createFeatureImages.sh};  creates the set of feature images given
  a set of co-registered input MRI from a single subject. 
  \item {\tt createModel.R}:  creates random forest model given the input csv 
  file of the feature image file names for all training data and set of truth
  label maps.
  \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
  (`{\tt MeanDecreaseAccuracy}') of each feature variable used 
  in constructing the model.
\end{itemize}
We also include a fully functional 2-D example which performs
both testing and training on sample challenge data.

@@

Brain tumor image data used in this work were obtained from the NCI-MICCAI 2013 
Challenge on Multimodal Brain Tumor Segmentation%
\footnote{
http://martinos.org/qtim/miccai2013/index.html
}
organized by K. Farahani, M. Reyes,B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. The challenge database contains fully anonymized images from the following institutions: 
ETH Zurich, University of Bern, University of Debrecen, and University of Utah and 
publicly available images from the Cancer Imaging Archive (TCIA).  Both training 
and testing data were made freely available through the Creative Commons Attribution-NonCommercial 3.0 license.

@@

\paragraph{Miscellaneous Feature Images}
Two additional feature images are calculated which seem to aid in 
classification as measured by the random forest variable importance
(relative importance plots are provided in the Results section).
First, the binary mask image is used to calculate the distance image 
\citep{maurer2003}.  Second, from the symmetric template deformation 
mapping we calculate the log Jacobian image.  Related to the earlier
distance feature image, we calculate the distance mask image of the symmetric
template and then warp that distance image to the individual subject.

@@

For use with the Challenge and Leaderboard data, cohort-specific models (both 
low-grade and 
high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
supplied training data.  Initial testing of our proposed framework was performed 
on the training data using a leave-one-out strategy.  Once the
feature images are created for each subject, the resulting images of the entire
training cohort are organized in a csv file for input into the R script
{\tt createModel.R}.  Other possible input parameters include the requested 
number of trees, number of samples per label, and number of threads for parallel
processing.  The output is an {\tt RData} file describing the random forest
model which can be used for future predictions.

Additionally, the {\tt randomForest} package provides a couple measurements 
for determining the importance of chosen features when constructing the model.  
This aids in potential feature pruning or intuiting model behavior.  As mentioned
earlier, we provide the R script {\tt plotVariableImportance.R} to render
one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
with the omission of a single feature or variable is tracked and averaged.  Thus,
those features which have the greatest decrease in mean accuracy are considered
to be the most discriminative.%
\footnote{
There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.  Correctives have been proposed but we ignore these issues for this particular application.
}

@@

However, as explained previously, the brain tumor segmentation 
methodology that we have made available is only a small part of the 
larger software package that we have created in ANTsR.  Not only does
ANTsR significantly facilitate the development of the work discussed,
but it provides an interface to one of the most powerful statistical
packages available in R.  The combination of the well-known ANTs software
package with R provides tremendous potential for future insightful analyses 
and graphical displays.

>>>>

\title{Multivariate Neuroimage Analysis with ANTsR: Application to Supervised Brain Tumor Segmentation using Random Forests}

@@

ANTs (Advanced Normalization Tools) originated with the open source availing
of high-performance registration algorithms for neuroimage analysis
\citep{avants2008a} built upon the mature Insight Toolkit (ITK)
of the National Institutes of Health.  Since then, ANTs has grown to include 
several algorithmic solutions for robust medical image analysis including bias 
correction \citep{tustison2010}, $n$-tissue multivariate segmentation 
\citep{avants2011}, template construction \citep{avants2010}, and cortical 
thickness estimation \citep{das2009} (many of which have been
introduced into ITK partially in an attempted leveraging of Linus's Law---``Given enough eyeballs, all bugs are shallow'').  
However, in the evolution of the toolkit, it became clear 
that robust statistical machinery was lacking for making inferences regarding
the data produced during the course of ANTs processing.  ANTsR was developed
specifically to provide an interface between ANTs, a 
powerful neuroimaging toolkit for producing reliable imaging data 
transformations, and the R project%
\footnote{
http://www.r-project.org
}
for statistical computing and visualization thus providing a complete
set of tools for multivariate neuroimage analysis.

%In addition to describing ANTsR basics and how it can be
%used generally in multivariate neuroimaging studies, 
We showcase ANTsR with
the particularly difficult application of supervised brain tumor
segmentation \citep{bauer2013} for which we use the increasingly popular 
random forests machine learning technique.  Although random forests have
been proposed previously in the literature for, among other things, 
supervised brain segmentation (e.g. \cite{geremia2011}), 
we demonstrate that ANTsR significantly facilitates the construction 
of a fully functional, parallelizable workflow for such a task and 
that the performance of our proposed framework has proven extremely effective in 
actual practice, specifically
 in the recent MICCAI 2013 Multimodal Brain Tumor Segmentation
Challenge (see Figure \ref{fig:award}).%
\footnote{
http://martinos.org/qtim/miccai2013/
}
\begin{figure}
  \centering
  \includegraphics[width=90mm]{Figures/award.pdf}
  \caption{First place certificate from the MICCAI 2013 BRATS challenge.}
  \label{fig:award}
\end{figure}

@@

Careful consideration of available options
led to the adoption of R to complement ANTs capabilities resulting in the
ANTsR package.
%The R project for statistical computing,%
%\footnote{
%www.r-project.org
%}
% or more compactly
%`R', is an environment for statistical computation
%and data visualization.  
R's open source code base
and add-on packages coupled with its rapidly growing 
community of developers and users has caused wide-scale
adoption within both academia and industry.

@@

As mentioned earlier, we have made this entire framework
available as open source.  In addition to the ANTsR repository
already on github which houses both ANTs and ANTsR functionality, 
we created a special github repository specifically for this work
containing figures, references, and text.%
\footnote{
https://github.com/ntustison/ANTsAndArboles
}
Also, we posted all
scripts (R, shell, and perl) used to coordinate the ANTsR processing 
including:
\begin{itemize}
  \item {\tt applyModel.R}:  applies a random forest model to a new 
  feature data set from a testing subject resulting in a set of probability
  images (one for each label).
  \item {\tt applyTumorSegmentationModel.sh}:  generates the new feature image set 
  from the testing MRI (by calling {\tt createFeatureImages.sh}).
  organizes the file names in a csv file (via {\tt createCSVFileFromModel.R}),
  and applies the random forest model using {\tt applyModel.R}. 
  \item {\tt applyTumorSegmentationModelForCohort.pl}:  Coordinates tumor 
  segmentation on the computational cluster for a given cohort.
  \item {\tt createCSVFileFromModel.R}:  organizes the set of feature image
  file names in a csv file for input into {\tt applyModel.R}.
  \item {\tt createFeatureImages.sh};  creates the set of feature images given
  a set of co-registered input MRI from a single subject. 
  \item {\tt createModel.R}:  creates a random forest model given the input csv 
  file of the feature image file names for all training data and set of truth
  label maps.
  \item {\tt plotVariableImportance.R}:  produces a plot of the importance 
  ({\tt MeanDecreaseAccuracy}) of each feature variable used 
  in constructing the model.
\end{itemize}
We also include a fully functional 2-D example which performs
both testing and training on sample challenge data.

@@

Brain tumor image data used in this work were obtained from the NCI-MICCAI 2013 
Challenge on Multimodal Brain Tumor Segmentation%
\footnote{
http://martinos.org/qtim/miccai2013/index.html
}
organized by K. Farahani, M. Reyes, B. Menze, E. Gerstner, J. Kirby and J. Kalpathy-Cramer. 
The challenge database contains fully anonymized images from the following institutions: 
ETH Zurich, University of Bern, University of Debrecen, and University of Utah and 
publicly available images from the Cancer Imaging Archive (TCIA).  Both training 
and testing data were made freely available through the Creative Commons Attribution-NonCommercial 3.0 license.

@@

\paragraph{Miscellaneous Feature Images}
Three additional feature images are calculated which seem to aid in 
classification as measured by the random forest variable importance
(relative importance plots are provided in the Results section).
First, the binary mask image is used to calculate the distance image 
\citep{maurer2003}.  Second, from the symmetric template deformation 
mapping we calculate the log Jacobian image.  Related to the earlier
distance feature image, we calculate the distance mask image of the symmetric
template and then warp that distance image to the individual subject
for a third feature image.

@@

For use with the Challenge and Leaderboard data, cohort-specific models (both 
low-grade and 
high-grade glioma) for both GMM and MAP-MRF stages were created using only the 
supplied training data.  Prior to training, we segmented normal brain tissue \cite{avants2011}
for each data set by segmenting only the T1 image.  This was only to yield
a rough estimate of normal brain tissue to augment the already provided 
pathology labels.  This resulted in seven labels (csf, gray matter, white matter,
necrosis, edema, enhancing, and non-enhancing tumor)
characterizing each brain.

Initial testing of our proposed framework was performed 
on the training data using a leave-one-out strategy.  Once the
feature images are created for each subject, the resulting images of the entire
training cohort are organized in a csv file for input into the R script
{\tt createModel.R}.  Other possible input parameters include the requested 
number of trees, number of samples per label, and number of threads for parallel
processing.  The output is an {\tt RData} file describing the random forest
model which can be used for future predictions.

Additionally, the {\tt randomForest} package provides  measurements 
for determining the importance of chosen features when constructing the model.  
This aids in potential feature pruning or intuiting model behavior.  As mentioned
earlier, we provide the R script {\tt plotVariableImportance.R} to render
one such quantity denoted as {\tt MeanDecreaseAccuracy}.  During model construction
(specifically the out-of-bag error calculation stage), the decrease in prediction accuracy
with the omission of a single feature or variable is tracked and averaged.  Thus,
those features which have the greatest decrease in mean accuracy are considered
to be the most discriminative.%
\footnote{
There is an issue with consistently ranking correlated predictors as described at {\tt http://www.r-bloggers.com/random-forest-variable-importance/} since the permutation testing performed for predictive accuracy assessment assumes predictor independence.  Correctives have been proposed but we ignore these issues for this particular application.
}

@@

However, as explained previously, the brain tumor segmentation 
methodology that we have made available is only a small part of the 
larger software package that we have created in ANTsR.  Not only does
ANTsR significantly facilitate the development of the work discussed,
but it provides an interface to one of the most powerful statistical
packages available in R.  The combination of the well-known ANTs software
package with R provides tremendous potential for future insightful analysis.
